The watershed transform in ITK - discussion and new developments

### 2006-06-08 article/Article.tex
try to clean and complete the article
darcs-hash:20060608151103-2fc9d-3e0667377f66e4c80494342e372b5ee89fb24101.gz

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images. These
definitions rigorously defined the behavior in the presence of
plateaus and lead to some new algorithms based on graph theory. Two
examples of these algorithms are {\em hill-climbing} and {\em image
integration} and will be discussed in more detail in this report.
Figures \ref{fig:cthead1-markers}, \ref{fig:circle} and \ref{fig:difficult-cases} illustrate
the possible problems with plateaus.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=0.7]{cthead1}}
\subfigure[User defined labels colored with LabelOverlayImageFilter]{\includegraphics[scale=0.7]{cthead1-markers-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mf-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mF-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1Mf-rgb}}
\subfigure[Morphological watershed]{\includegraphics[scale=0.7]{cthead1MF-rgb}}
\subfigure[ITK watershed]{\includegraphics[scale=0.7]{cthead1itk-rgb}}
\caption{Example of plateau problem. The blue marker is clearly extended where it should not with the ITK watershed.\label{fig:cthead1-markers}}
\end{center}
\end{figure}

@@

\subsection{The topological watershed}
An alternative approach, known as the topological watershed, has
recently been developed \cite{topologicalWatershed}. The topological watershed modifies
the topology of the surface so that all pixels that are not watershed
lines become regional minima and watershed lines are modified so that
they are more suitable for use in measuring saliency of boundaries
between regions. Algorithms implementing the topological watershed
also exploit priority queues, but don't appear to be able to avoid the
need for explicit lower completion or reconstruction by erosion.

@@

I don't understand the implications of the topological watershed yet.

\section{Rainfall model}
\label{sect:rainfall}
{\em where should we put this?}  The original concept behind the
watershed transform was rain falling on a terrain and flowing down
paths of steepest descent to local minima. There has been a lot of
work on alternatives to this approach because of the problems with
it. These problems are basically to do with flat zones in an image
which have no paths of steepest descent leading from them, which means
that water would just sit in the one spot. A simple minded
implementation of this idea would therefore produce a separate
catchment basin for each pixel in the flat zone, which is not a very
useful result. The ITK watershed uses a number of steps to avoid this problem:
\begin{enumerate}
\item Label all minima and flat zones.
\item Trace all unlabelled pixels to a labelled one using gradient descent.
\item Check each flat region and find the lowest neighbor. Change the label of the flat region to that of the lowest neighbor.
\end{enumerate}

@@

Using ordered queues.

This algorithm does not mark the watershed lines.

The queue only contains the borders of the growing regions.

@@

@@

Also using ordered queues.

@@

This algorithm mark the watershed lines.

The queue only contains the borders of the growing regions.

It should be noted that this algorithm never propagates watershed
lines, and that the watershed lines it produces have a width of 1 pixel.

@@

The difference between algorithms 2 and 3 is the order in which
labelling and queueing happens.

{\em a summary of our findings comparing the last 2 algorithms}

The new work implements algorithms 2 and 3 -- algorithm 2 is used when
the watershed lines are not marked while algorithm 3 is used when they
are.

@@

\caption{Three difficult test images. The first image, plateau, illustrates a 
bad result produced by ITK's watershed in which most of the area is
assigned to one minima. This problem is also shown in Figure
\ref{fig:cthead1-markers}. The morphological watershed give a better
result, but not yet perfect: the separation line should not be curved
as it is. This problem is also visible in Figure
\ref{fig:embryo-mws}. The reason for the problem is as follows - the 
example is attempting to split a plateau into two regions, and should
therefore produce a Voronoi tesselation, with the two markers being
separated by a straight line. This does not happen because various
approximations have been made, as mentioned in Section
\ref{sect:hillclimbing}, that influence the estimation of the distance
function. Even if a more accurate underlying distance function was
used the computations are still being carried out on a grid and so are
limited by the digital representation of lines.
The next image, thick line, again show that ITK watershed can produce
strange results with a green label which is where we should have the
light blue one. Morphological watershed gives good results. The last
image, button hole, is a very difficult case for watershed algorithms
which mark watershed lines: the watershed pixel at the bottom of the
image stop the label propagation. There are several possible
behaviors: produce a large watershed line which covers all the zone
(Vincent-Soille algorithm, not shown here), let the label on the top
of the image propagate into the zone (Meyer algorithm), split the zone
in 2 zones with the 2 labels of the bottom of the image (topological
watershed, not shown here).\label{fig:difficult-cases}}
\end{center}
\end{figure}

\section{Finding regional minima}
The marker based watershed filter can be used as a convention
watershed algorithm by finding the regional minima and using them as
markers. Regional minima (and maxima) are often useful, and there are
a number of different ways of finding them:
\begin{itemize}
\item Using reconstruction approaches mentioned earlier - 
the problem with these is that a value of 1 is subtracted from the
original to produce the image from which the reconstruction is carried
out, which makes the approach less useful for floating point
types. The advantage of this approach is that it can be easily
extended to find {\em image dynamics}, which are often more stable and
less subject to noise than regional minima.
\item The way ITK does it, which is related to the connected 
component labeller. Finds all flat regions and then figures out which
are plateaus, minima and maxima by looking at the boundary values.
\item Flooding transform (not sure where this came from) which 
sets anything not a regional minima to a specified max value. The
algorithm assumes that the border is set to the maximum value. Pixels
are visited in raster order. If a pixel $A$ has a neighbour $B$ and
$P(A) > p(B)$ (i.e grey level of A is greater than B) then all pixels
connected to $A$ with value $P(A)$ are set to the maximum value. This
is a recursive algorithm that finds pixels that cannot be a regional
minimum (because they have a lower neighbour) and then eliminates any
connected region that also cannot be a regional minima.

The result would then need to be thresholded and labelled to produce
markers. This approach is not limited by the pixel type, but does
require a more random visit order when flooding, which could restrict
performance on larger images.
\end{itemize}

@@

The watershed from markers can be implemented with the ITK watershed,
by using the technique of {\em minima imposition} exposed in Section
\ref{sect:ws-selected-minima}. Several steps are required:
\begin{itemize}
  \item Impose the minima to the input image, with the
\verb$MinimaImpositionImageFilter$
  \item Perform the watershed with the \verb$WatershedImageFilter$
  \item Restore the labels produced by the \verb$WatershedImageFilter$.
\verb$WatershedImageFilter$ produce arbitrary labels which doesn't match
the labels in the marker image. There is currently no filter able to
find the label correspondance in ITK, so this step must be performed by
hand with a code like the one before. We can use a
\verb$ChangeLabelImageFilter$ to change the labels.
\small \begin{verbatim}
filter->Update();

@@

This approach has several drawbacks:
\begin{itemize}
  \item it doesn't work if there some markers not separated by background
pixels - they are considered as a single marker
  \item restoring the labels is not included in a filter and so break the
ITK pipeline model
  \item it's less efficient, as shown in Section \ref{sect:perf}
  \item it gives poor results with plateaus
\end{itemize}

@@

{\em Don't think this should be here. Is there any compiler that will
represent bool images as bit vectors? It might save memory but will
increase access time}.  We prefered using a separated binary image (of
type \verb$itk::Image<bool, dim>$) to store this information. This
solution avoid the need to copy the image at the end, and so should be
more efficient in computation time, and require only one bit per pixel
and so should be more memory efficient as well.

@@

\verb$MorphologicalWatershedImageFilter$ performs a watershed transformation
without markers. It encapsulate several filters:
\begin{itemize}
  \item \verb$HMinimaImageFilter$
  \item \verb$RegionalMinimaImageFilter$
  \item \verb$ConnectedComponentImageFilter$
  \item \verb$MorphologicalWatershedFromMarkersImageFilter$
\end{itemize}
While \verb$RegionalMinimaImageFilter$, \verb$ConnectedComponentImageFilter$,
and \verb$MorphologicalWatershedFromMarkersImageFilter$ are the usal steps
of the watershed usage, the \verb$HMinimaImageFilter$ is a convenient way
to avoid the classical watershed's over segmentation by allowing the user
to select the dynamic of the minima. The minimum dynamic of minima can be
selected with \verb$SetLevel()$.

@@

The binary image produced by the \verb$ThresholdImageFilter$ contains lot
of small objects. To remove them, we use an {\em opening by reconstruction}
which have the advantage to preserve the egdes of the object.

@@

There is now only one binary object in the image, but it contains lots of holes
which will produce an unsable distance map for the watershed. We fill them
with a {\em hole filling} filter. This algorithm, based on a morphological
reconstruction, fill the hole in a binary object but preserve the edges of the
object. In this case, it is important to keep the two cavities between the two
pronucleus unchanged.

@@

Then we compute the distance map which will be used by the watershed. The
binary object must be inverted to compute the distance {\em inside} the
object, and the distance map must be inverted to used by the watershed.

@@

The watershed is then performed to split the two pronucleus. A level
is used to avoid the too small minima in the border of the object and thus
avoid over segmentation. The label produced by the watershed are then
displayed on top of the original image.

The watershed extends the labels on the entire image. We use a
\verb$MaskImageFilter$ to limit the labels to the object. Another commonly
used method is to use the background as a marker and to perform a watershed
from markers.

@@

The same watershed is done with the ITK watershed filter\footnote{A
$RelabelComponentImageFilter$ is used here because unsigned long is
not available in python for the $MaskImageFilter$}.

@@

\section{Related work}
This work have lead us to develop some other filters to:
\begin{itemize}
  \item Find regional extrema \cite{regionalExtrema}
  \item Superimpose a colored label image onto an image \cite{labelOverlay}
  \item Impose the minima \cite{minimaImposition}
  \item Label the connected component \cite{connectedComponent}
  \item Invert the intensity of an image \cite{invert}
  \item Enhance the morphological reconstruction \cite{recons}
  \item Compute some informations about the shape of the labeled objects (not published yet)
\end{itemize}

>>>>

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images\footnote{A lower complete
image is an image where there is no plateaus, excepted plateaus
which are regional minima. A common way to compute the lower completion of an
image, is to add the distance from the nearest lower pixel to the pixels of all
the plateaus. Lower completion have the property to produce the same image than
the input image if the input image is already lower complete. By lower completing
an image, we introduce a new order relation in the plateaus pixels which can be
used to solve the plateaus problem with watershed.}. These
definitions rigorously defined the behavior in the presence of
plateaus and lead to some new algorithms based on graph theory. Two
examples of these algorithms are {\em hill-climbing} and {\em image
integration} and will be discussed in more detail in this report.
Figures \ref{fig:cthead1-markers}, \ref{fig:circle} and \ref{fig:difficult-cases} illustrate
the possible problems with plateaus.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=0.7]{cthead1}}
\subfigure[User defined labels]{\includegraphics[scale=0.7]{cthead1-markers-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mf-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mF-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1Mf-rgb}}
\subfigure[Morphological watershed]{\includegraphics[scale=0.7]{cthead1MF-rgb}}
\subfigure[ITK watershed]{\includegraphics[scale=0.7]{cthead1itk-rgb}}
\caption{Example of plateau problem. The blue marker is clearly extended where it should not with the ITK watershed. Labels are colorized and superimposed on the input image.\label{fig:cthead1-markers}}
\end{center}
\end{figure}

@@

% \subsection{The topological watershed}
An alternative approach, known as the topological watershed, has
recently been developed \cite{topologicalWatershed}. The topological watershed modifies
the topology of the surface so that all pixels that are not watershed
lines become regional minima and watershed lines are modified so that
they are more suitable for use in measuring saliency of boundaries
between regions. Algorithms implementing the topological watershed
also exploit priority queues, but don't appear to be able to avoid the
need for explicit lower completion or reconstruction by erosion.

@@

% I don't understand the implications of the topological watershed yet.

% \section{Rainfall model}
\label{sect:rainfall}
The original concept behind the
watershed transform was rain falling on a terrain and flowing down
paths of steepest descent to local minima. There has been a lot of
work on alternatives to this approach because of the problems with
it. These problems are basically to do with flat zones in an image
which have no paths of steepest descent leading from them, which means
that water would just sit in the one spot. A simple minded
implementation of this idea would therefore produce a separate
catchment basin for each pixel in the flat zone, which is not a very
useful result. The ITK watershed uses a number of steps to avoid this problem:
\begin{enumerate}
\item Label all minima and flat zones.
\item Trace all unlabelled pixels to a labelled one using gradient descent.
\item Check each flat region and find the lowest neighbor. Change the label of the flat region to that of the lowest neighbor.
\end{enumerate}

@@

@@

It should be noticed that:
\begin{itemize}
  \item this algorithm use ordered queues,
  \item this algorithm doesn't mark the watershed lines,
  \item the ordered queue only contains the borders of the growing regions.
\end{itemize}

@@

@@

It should be noticed that:
\begin{itemize}
  \item this algorithm mark the watershed lines,
  \item the queue only contains the borders of the growing regions,
  \item this algorithm never propagates watershed
lines, and that the watershed lines it produces have a width of 1 pixel.
Also, the watershed lines form a connected obejct if we consider the
opposite connectivity than the one used in the algorithm: with a 2D
image the watershed lines are 8-connected if the algorithm is run
as 4-connected, and 4-connected if it is run as 8-connected.
\end{itemize}

@@

% The difference between algorithms 2 and 3 is the order in which
% labelling and queueing happens.
% 
% {\em a summary of our findings comparing the last 2 algorithms}
% 
% The new work implements algorithms 2 and 3 -- algorithm 2 is used when
% the watershed lines are not marked while algorithm 3 is used when they
% are.

@@

\caption{Three difficult test images. The first image, plateau, illustrates a 
bad result produced by ITK's watershed in which most of the area is
assigned to one minima. This problem is also shown in Figure
\ref{fig:cthead1-markers}. The morphological watershed give a better
result, but not yet perfect: the separation line should not be curved
as it is. This problem is also visible in Figure
\ref{fig:embryo-mws}. The reason for the problem is as follows - the 
example is attempting to split a plateau into two regions, and should
therefore produce a Voronoi tesselation, with the two markers being
separated by a straight line. This does not happen because various
approximations have been made, as mentioned in Section
\ref{sect:hillclimbing}, that influence the estimation of the distance
function. Even if a more accurate underlying distance function was
used the computations are still being carried out on a grid and so are
limited by the digital representation of lines.
The next image, thick line, again show that ITK watershed can produce
strange results with a green label which is where we should have the
light blue one. Morphological watershed gives good results. The last
image, button hole, is a very difficult case for watershed algorithms
which mark watershed lines: the watershed pixel at the bottom of the
image stop the label propagation. There are several possible
behaviors: produce a large watershed line which covers all the zone
(Vincent-Soille algorithm, not shown here), let the label on the top
of the image propagate into the zone (Meyer algorithm), split the zone
in 2 zones with the 2 labels of the bottom of the image (topological
watershed, not shown here). See \cite{najman03watershed} for more
details.\label{fig:difficult-cases}}
\end{center}
\end{figure}

% \section{Finding regional minima}
% The marker based watershed filter can be used as a convention
% watershed algorithm by finding the regional minima and using them as
% markers. Regional minima (and maxima) are often useful, and there are
% a number of different ways of finding them:
% \begin{itemize}
% \item Using reconstruction approaches mentioned earlier - 
% the problem with these is that a value of 1 is subtracted from the
% original to produce the image from which the reconstruction is carried
% out, which makes the approach less useful for floating point
% types. The advantage of this approach is that it can be easily
% extended to find {\em image dynamics}, which are often more stable and
% less subject to noise than regional minima.
% \item The way ITK does it, which is related to the connected 
% component labeller. Finds all flat regions and then figures out which
% are plateaus, minima and maxima by looking at the boundary values.
% \item Flooding transform (not sure where this came from) which 
% sets anything not a regional minima to a specified max value. The
% algorithm assumes that the border is set to the maximum value. Pixels
% are visited in raster order. If a pixel $A$ has a neighbour $B$ and
% $P(A) > p(B)$ (i.e grey level of A is greater than B) then all pixels
% connected to $A$ with value $P(A)$ are set to the maximum value. This
% is a recursive algorithm that finds pixels that cannot be a regional
% minimum (because they have a lower neighbour) and then eliminates any
% connected region that also cannot be a regional minima.
% 
% The result would then need to be thresholded and labelled to produce
% markers. This approach is not limited by the pixel type, but does
% require a more random visit order when flooding, which could restrict
% performance on larger images.
% \end{itemize}

@@

The watershed from markers can be implemented with the ITK watershed,
by using the technique of {\em minima imposition} exposed in Section
\ref{sect:ws-selected-minima}. Several steps are required:
\begin{itemize}
  \item Impose the minima to the input image, with the
\verb$MinimaImpositionImageFilter$
  \item Perform the watershed with the \verb$WatershedImageFilter$
  \item Restore the labels produced by the \verb$WatershedImageFilter$.
\verb$WatershedImageFilter$ produce arbitrary labels which doesn't match
the labels in the marker image. There is currently no filter able to
find the label correspondance in ITK, so this step must be performed by
hand with a code like the one before. The \verb$ChangeLabelImageFilter$
is well designed for this task:
\small \begin{verbatim}
filter->Update();

@@

This approach has several drawbacks:
\begin{itemize}
  \item it doesn't work if there some markers are not separated by background
pixels - they are considered as a single marker
  \item restoring the labels is not included in a filter and so break the
ITK pipeline model
  \item it's less efficient, as shown in Section \ref{sect:perf}
  \item it still gives poor results with plateaus. A lower completion
sould be performed before to also solve this problem\footnote{Lower
completion is not implemented yet in ITK}.
\end{itemize}

@@

% {\em Don't think this should be here. Is there any compiler that will
% represent bool images as bit vectors? It might save memory but will
% increase access time}.  We prefered using a separated binary image (of
% type \verb$itk::Image<bool, dim>$) to store this information. This
% solution avoid the need to copy the image at the end, and so should be
% more efficient in computation time, and require only one bit per pixel
% and so should be more memory efficient as well.

@@

\verb$MorphologicalWatershedImageFilter$ performs a watershed transformation
{\em without} markers. It encapsulate several filters:
\begin{itemize}
  \item \verb$HMinimaImageFilter$
  \item \verb$RegionalMinimaImageFilter$
  \item \verb$ConnectedComponentImageFilter$
  \item \verb$MorphologicalWatershedFromMarkersImageFilter$
\end{itemize}
While \verb$RegionalMinimaImageFilter$, \verb$ConnectedComponentImageFilter$,
and \verb$MorphologicalWatershedFromMarkersImageFilter$ are the usal steps
of the watershed usage, the \verb$HMinimaImageFilter$ is a convenient way
to avoid the classical watershed's over segmentation by allowing the user
to select the dynamic of the minima. The minimum dynamic of minima can be
selected with \verb$SetLevel()$.

@@

The binary image produced by the \verb$ThresholdImageFilter$ contains lot
of small objects. To remove them, we use an {\em opening by reconstruction}
which has the advantage to preserve the egdes of the objects.

@@

There is now only one binary object in the image, but it contains lots of holes
which will produce an unusable distance map for the watershed. We fill them
with a {\em hole filling} filter. This algorithm, based on a morphological
reconstruction, fill the hole in a binary object but preserve the edges of the
object. In this case, it is important to keep the two cavities between the two
pronucleus unchanged.

@@

Then we compute the distance map which will be used by the watershed. The
binary object must be inverted to compute the distance {\em inside} the
object, and the distance map must be inverted to be used by the watershed.

@@

The watershed is then performed to split the two pronucleus. A \verb$Level$
is set to avoid the too small minima in the border of the object and thus
avoid over segmentation. The label produced by the watershed are then
displayed on top of the original image.

The watershed extends the labels on the entire image. We use a
\verb$MaskImageFilter$ to keep the labels inside the object. Another commonly
used method to keep the labeled zone in the object is to use the regional
minima {\em and} the background as markers, to perform a marker based watershed.

@@

The same watershed is done with the ITK watershed filter\footnote{A
$RelabelComponentImageFilter$ is used here because $unsigned long$ is
not one of the available types for the $MaskImageFilter$ in python}.

@@

\section{Related work}
This work has lead us to develop some other filters to:
\begin{itemize}
  \item Find regional extrema \cite{regionalExtrema}
  \item Superimpose a colored label image onto an image \cite{labelOverlay}
  \item Impose the minima \cite{minimaImposition}
  \item Label the connected component \cite{connectedComponent}
  \item Invert the intensity of an image \cite{invert}
  \item Enhance the morphological reconstruction \cite{recons}
  \item Compute some informations about the shape of the labeled objects (not published yet)
\end{itemize}
